Supermartingales and Termination Analysis of Statistical PCF

### 2021-01-15 LICS21/antitone-proofs.tex
Cut another proof

% !TEX root = main.tex
\label{apx:antitone}
\AntitonePartial*
\begin{proof}
  Take a closed term $M$ and an antitone sparse ranking function $f$ on $M$, with a corresponding antitone function $\epsilon$. Assume wlog that $\epsilon$ is convex (as if it isn't, we can just take its convex hull instead). As in Theorem \ref{thm:partial implies rankable}, define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
  \begin{align*}
    f_1(N) &= f(N) \text{ whenever $f(N)$ is defined},\\
    f_1(V) &= 0 \text{ for values $V$ not in the domain of $f$.}
  \end{align*}

>>>>

% !TEX root = main.tex
\label{apx:antitone}
\ACStrict*
\begin{proof}
First, as $(Y_n)$ is a supermartingale, $\expect{Y_n} \leq \expect{Y_0}$. 
Therefore
\begin{calculation}
\expect{Y_n \mid T > n}
   \step[=]{rearranging terms}
% \dfrac{\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
% + 
% \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
% - 
% \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
\dfrac{
\begin{array}{l}
\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
+ 
\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
\\
{} - 
\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}
\end{array}
}{\mathbb P[T > n]}
  \step[=]{definition of conditional expectation}
\dfrac{\expect{Y_n} - \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
  \step[\leq]{}
\dfrac{\expect{Y_n}}{\mathbb P[T > n]}
%  \step[\leq]{}
%\dfrac{\expect{Y_0}}{\mathbb P[T > n]}
\mathbin{=} \dfrac{\expect{Y_0}}{\mathbb P[T > n]}
\end{calculation}

\emph{Claim}: For all $0 < x \leq 1$, 
\(
\mathbb P[T > B_x] \leq x,
\)
where $B_x = \left \lceil \dfrac{\expect{Y_0}+1}{x \; \epsilon(\expect{Y_0} \, x^{-1})} \right \rceil$ and $\epsilon : \nnReal \to \pReal$ is the antitone function.

\smallskip

As the convex hull of $\epsilon$ (the greatest convex function less than or equal to it) satisfies all the conditions assumed of $\epsilon$, in addition to being convex, assume wlog that $\epsilon$ is convex.

Assume for a contradiction that $\mathbb P[T > B_x] > x$.
Then, take $n \leq B_x$. 
We have
%\lo{Change of justification of the second equality below: event, not discrete r.v. Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
\begin{calculation}
\expect{Y_n-Y_{n+1}}
%  \step[=]{}
\mathbin{=}
\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
  \step[\geq]{}
\expect{\epsilon(Y_n) \cdot {\bf 1}_{\set{T > n}}}
%  \step[=]{definition of conditional expectation for a discrete variable}
%\step[=]{}
\mathbin{=}
\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
  \step[\geq]{Jensen's inequality}
\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
  \step[\geq]{proved earlier}
\mathbb P[T > n]\ \epsilon\Big(\dfrac{\expect{Y_0}}{\mathbb P[T > n]}\Big)
  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
x \; \epsilon\big(\expect{Y_0} \, x^{-1}\big).
\end{calculation}
Therefore, by a telescoping sum
\begin{align*}
\expect{Y_{B_x}} 
&= \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})\\
&\leq -1 < 0 
\end{align*}
%\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
which is a contradiction, therefore the claim must be true, therefore $P[T > n] \to 0$ as $n \to \infty$, therefore $T < \infty$ a.s.
\end{proof}

\AntitonePartial*
\begin{proof}
  Take a closed term $M$ and an antitone sparse ranking function $f$ on $M$, with a corresponding antitone function $\epsilon$. Assume wlog that $\epsilon$ is convex (as if it isn't, we can just take its convex hull instead). As in Theorem \ref{thm:partial implies rankable}, define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
  \begin{align*}
    f_1(N) &= f(N) \text{ whenever $f(N)$ is defined},\\
    f_1(V) &= 0 \text{ for values $V$ not in the domain of $f$.}
  \end{align*}

### 2021-01-15 LICS21/antitone.tex
Introduce Conjecture environment for the completeness conjecture.
Fix typo (g_1 vs g_2)
Cut another proof

\begin{theorem}
\label{thm:a-c strict}
Let $(Y_n)_{n \geq 0}$ be an antitone strict supermartingale w.r.t.~stopping time $T$. 
Then $T < \infty$ a.s.
\end{theorem}

\begin{proof}
First, as $(Y_n)$ is a supermartingale, $\expect{Y_n} \leq \expect{Y_0}$. 
Therefore
\begin{calculation}
\expect{Y_n \mid T > n}
   \step[=]{rearranging terms}
% \dfrac{\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
% + 
% \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
% - 
% \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
\dfrac{
\begin{array}{l}
\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
+ 
\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
\\
{} - 
\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}
\end{array}
}{\mathbb P[T > n]}
  \step[=]{definition of conditional expectation}
\dfrac{\expect{Y_n} - \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
  \step[\leq]{}
\dfrac{\expect{Y_n}}{\mathbb P[T > n]}
%  \step[\leq]{}
%\dfrac{\expect{Y_0}}{\mathbb P[T > n]}
\mathbin{=} \dfrac{\expect{Y_0}}{\mathbb P[T > n]}
\end{calculation}

\emph{Claim}: For all $0 < x \leq 1$, 
\(
\mathbb P[T > B_x] \leq x,
\)
where $B_x = \left \lceil \dfrac{\expect{Y_0}+1}{x \; \epsilon(\expect{Y_0} \, x^{-1})} \right \rceil$ and $\epsilon : \nnReal \to \pReal$ is the antitone function.

\smallskip

As the convex hull of $\epsilon$ (the greatest convex function less than or equal to it) satisfies all the conditions assumed of $\epsilon$, in addition to being convex, assume wlog that $\epsilon$ is convex.

Assume for a contradiction that $\mathbb P[T > B_x] > x$.
Then, take $n \leq B_x$. 
We have
%\lo{Change of justification of the second equality below: event, not discrete r.v. Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
\begin{calculation}
\expect{Y_n-Y_{n+1}}
%  \step[=]{}
\mathbin{=}
\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
  \step[\geq]{}
\expect{\epsilon(Y_n) \cdot {\bf 1}_{\set{T > n}}}
%  \step[=]{definition of conditional expectation for a discrete variable}
%\step[=]{}
\mathbin{=}
\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
  \step[\geq]{Jensen's inequality}
\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
  \step[\geq]{proved earlier}
\mathbb P[T > n]\ \epsilon\Big(\dfrac{\expect{Y_0}}{\mathbb P[T > n]}\Big)
  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
x \; \epsilon\big(\expect{Y_0} \, x^{-1}\big).
\end{calculation}
Therefore, by a telescoping sum
\begin{align*}
\expect{Y_{B_x}} 
&= \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})\\
&\leq -1 < 0 
\end{align*}
%\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
which is a contradiction, therefore the claim must be true, therefore $P[T > n] \to 0$ as $n \to \infty$, therefore $T < \infty$ a.s.
\end{proof}

Note that it is essential in this proof that $\epsilon$ be defined for all $\mathbb R_{\geq 0}$, not just the values that $(Y_n)_{n \geq 0}$ actually takes (or at least if $(Y_n)_{n \geq 0}$ is uniformly bounded, $\epsilon$ must be positive at the supremum as well as the actual values).

@@

\begin{example}[Unbiased random walk]
\label{ex:unbiased random walk}
For $M_2$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \ln(x+1) + 1$.
Using shorthand 
$\Theta_2 = \tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
we can define an antitone sparse ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ 
%with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
by
\(
f_2: {\Theta_2} \, n \mapsto 
g_2(n)
\)
and
$0 \mapsto 0$
for $n \in \mathbb N$.
% \begin{align*}
% {\Theta_2} \, n 
% &\mapsto 
% g_2(n)
% \\
% 0 &\mapsto 0.
% \end{align*}
For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$,\footnote{Technically this is not quite correct, because of the distinction between $\underline n + 1$ and $\underline{n + 1}$, but this ranking function can still be made to work by appealing to \Cref{thm:confluent ranking}, as can the others in this section with similar issues.} each with probability $1/2$, with one $\tY$-reduction.
Now
\begin{align*}%\textstyle
& g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
  &=  \ln \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) 
  =  \frac 1 2 \ln\left(\frac{(n+1)^2}{n(n+2)}\right) \\
  &=  \frac 1 2 \ln\left(1 + \frac 1 {n(n+2)}\right) 
  >  \frac 1 {n(n+2) + 1} \\
  &>  \frac 1 2 \frac 1 {(n+1)^2} + \frac 1 2 \frac 1 {(n+3)^2}
\end{align*}
Moreover $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
Therefore by setting $\epsilon_2(x) = \frac 1 {(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met. It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone sparse ranking function, and $M_2$ is AST.
\end{example}

>>>>

\begin{restatable}{theorem}{ACStrict}
\label{thm:a-c strict}
Let $(Y_n)_{n \geq 0}$ be an antitone strict supermartingale w.r.t.~stopping time $T$. 
Then $T < \infty$ a.s.
\end{restatable}

It is essential in this theorem that $\epsilon$ be defined for all $\mathbb R_{\geq 0}$, not just the values that $(Y_n)_{n \geq 0}$ actually takes (or at least if $(Y_n)_{n \geq 0}$ is uniformly bounded, $\epsilon$ must be positive at the supremum as well as the actual values).

@@

\begin{example}[Unbiased random walk]
\label{ex:unbiased random walk}
For $M_2$, define the function $g_2 : \nnReal \to \nnReal$ by $g_2(x) = \ln(x+1) + 1$.
Using shorthand 
$\Theta_2 = \tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
we can define an antitone sparse ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ 
%with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
by
\(
f_2: {\Theta_2} \, n \mapsto 
g_2(n)
\)
and
$0 \mapsto 0$
for $n \in \mathbb N$.
% \begin{align*}
% {\Theta_2} \, n 
% &\mapsto 
% g_2(n)
% \\
% 0 &\mapsto 0.
% \end{align*}
For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$,\footnote{Technically this is not quite correct, because of the distinction between $\underline n + 1$ and $\underline{n + 1}$, but this ranking function can still be made to work by appealing to \Cref{thm:confluent ranking}, as can the others in this section with similar issues.} each with probability $1/2$, with one $\tY$-reduction.
Now
\begin{align*}%\textstyle
& g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
  &=  \ln \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) 
  =  \frac 1 2 \ln\left(\frac{(n+1)^2}{n(n+2)}\right) \\
  &=  \frac 1 2 \ln\left(1 + \frac 1 {n(n+2)}\right) 
  >  \frac 1 {n(n+2) + 1} \\
  &>  \frac 1 2 \frac 1 {(n+1)^2} + \frac 1 2 \frac 1 {(n+3)^2}
\end{align*}
Moreover $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
Therefore by setting $\epsilon_2(x) = \frac 1 {(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met. It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone sparse ranking function, and $M_2$ is AST.
\end{example}

### 2021-01-15 LICS21/applications.tex
Cut another proof

% !TEX root = main.tex
\section{Applications}
\label{sec:applications}
Although the definition of potential positions in a term in \Cref{sec:confluent} was intended simply to define a variant of sampling semantics that has a restricted version of the Church-Rosser property, it is likely to also be applicable to probabilistic inference algorithms as well.

In the Metropolis-Hastings inference algorithm, the program is executed, with a random value being chosen each time a $\tsample$ redex is evaluated. The trace of samples used is recorded, then for the next step, it is modified slightly before the program is run again. This time, when $\tsample$ redexes are evaluated, they may be taken from the pre-determined trace instead of chosen randomly. The change to the trace may be accepted or rejected, depending on the probability weight associated with each execution of the program (based on a conditioning language construct), and after this process is repeated sufficiently many times, the distribution of results of the program tends towards the true posterior distribution.

@@

It is possible to reduce this problem by labelling the samples in the trace by something other than the order in which they are used. If the labels of samples in the trace correspond closely to the roles that the samples play in the program execution, this will increase the efficiency of the inference algorithm. 
In \cite{wingate2011lightweight}, for example, something rather like a stack trace is used, although this labelling scheme turns out to have various problems \cite{kiselyov2016problems}. 
Potential positions could also be used for a similar purpose, although in practice, a more compact representation than an entire skeletal reduction sequence would be necessary. 
Selecting an appropriate addressing scheme can also be important in the context of variational inference \cite[\S 6.2.2]{paige2016automatic}.

@@

\medskip
Another possible application for the confluent trace semantics would be as a justification, within a sampling-style semantics, of some forms of equational reasoning. By introducing reduction steps that go backwards to terms not reachable from the original term, it may in some cases be simplified, or equivalences between terms may be proved, by appealing to confluence results such as \Cref{churchRosser} and \Cref{thm:CbvIsTerminatingest}.

>>>>

% !TEX root = main.tex
\section{Applications}
\label{sec:applications}
Although the definition of potential positions in a term in \Cref{sec:confluent} was intended simply to define a variant of trace semantics that has a restricted version of the Church-Rosser property, it is likely to also be applicable to inference algorithms (for probabilistic programs) as well.

%\lo{What you described in the following paragraph is not the Metropolis-Hastings algorithm (for MH algorithm, see e.g.~Algorithm 1, Slide 18 of 5-MCMC-part2.pdf \footnote{\url{https://www.cs.ox.ac.uk/teaching/materials20-21/SPP/Lectures/5-MCMC-part2.pdf}, \cite{Ong20}}), but a version of component-wise MH adapted to (models \emph{qua}) probabilistic programs (see Slide 36 onwards, 6-operational semantics.pdf \footnote{\url{https://www.cs.ox.ac.uk/teaching/materials20-21/SPP/Lectures/6-Operational-Semantics-21Nov20.pdf}, \cite{Ong20} \label{fnote:lmh}}).}

In lightweight Metropolis-Hastings (LMH) or single-site MH (originally suggested by \cite{wingate2011lightweight}; see also \cite{rainforth2017Automating}),
the program is executed, with a random value being chosen each time a $\tsample$ redex is evaluated. The trace of samples used is recorded, then for the next step, it is modified slightly before the program is run again. This time, when $\tsample$ redexes are evaluated, they may be taken from the pre-determined trace instead of chosen randomly. The change to the trace may be accepted or rejected, depending on the probability weight associated with each execution of the program (based on a conditioning language construct), and after this process is repeated sufficiently many times, the distribution of results of the program tends towards the true posterior distribution.

@@

It is possible to reduce this problem by labelling the samples in the trace by something other than the order in which they are used. If the labels of samples in the trace correspond closely to the roles that the samples play in the program execution, this will increase the efficiency of the inference algorithm. 
In \cite{wingate2011lightweight}, for example, something rather like a stack trace is used, although this assdressing scheme turns out to have various problems \cite{kiselyov2016problems,Hur2015a}. 
Potential positions could also be used for a similar purpose, although in practice, a more compact representation than an entire skeletal reduction sequence would be necessary. 
%\lo{It would be very interesting (and rewarding, I believe) to work out the details. There is a formalisation of LMH for Idealised Anglican in \Cref{fnote:lmh}. SPCF is a much nicer probabilistic programming language (in purified / idealised form).}
Selecting an appropriate addressing scheme can also be important in the context of variational inference \cite[\S 6.2.2]{paige2016automatic}.

@@

\medskip

Another possible application for the confluent trace semantics would be as a justification, within a trace-style semantics, of some forms of equational reasoning. 
By introducing reduction steps that go backwards to terms not reachable from the original term, it may in some cases be simplified, or equivalences between terms may be proved, by appealing to confluence results such as \Cref{churchRosser} and \Cref{thm:CbvIsTerminatingest}.
\lo{Can you provide an example (and present it in the appendix)?} \akr{I don't have any compelling example for this in mind.}
